\mutagen: Coverage-Guided, Property-Based Testing using Exhaustive Structure-Preserving Mutations

### 2022-01-13 papers/ast22/main.tex
Added AST22 version

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf, anonymous, review]{acmart}

@@

%% \titlespacing*{\section}{0pt}{5pt}{5pt}
%% \titlespacing*{\subsection}{0pt}{5pt}{5pt}
%% \titlespacing*{\subsubsection}{0pt}{5pt}{5pt}
\titlespacing*{\paragraph}{10pt}{0pt}{10pt}

@@

%% These commands are for a JOURNAL article.
\acmConference[ICST 2022]{15th IEEE International Conference on Software
  Testing, Verification and Validation}{Apr 4–13, 2022}{}

@@

%% \author{Agust\'in Mista}
%% \email{mista@chalmers.se}
%% \affiliation{%
%%   \institution{Chalmers University of Technology}
%%   \city{Gothenburg}
%%   \country{Sweden}
%% }

%% \author{Alejandro Russo}
%% \email{russo@chalmers.se}
%% \affiliation{%
%%   \institution{Chalmers University of Technology}
%%   \city{Gothenburg}
%%   \country{Sweden}
%% }

@@

%% \vspace{-5pt}

@@

%% In this light, CGPT is likely to preserve the semantic structure of the mutated
%% data, as mutations are applied directly at the data type level, e.g., the
%% program's AST in case of generating code.
%
Moreover, CGPT relies on execution traces to distinguish which executed test
cases were interesting and are therefore worth mutating --- a popular technique
known as \emph{coverage-guided fuzzing} and popularized by tools like \emph{AFL}
\cite{afl}.
%
Mutated test cases are considered interesting for mutation only when they
produce new execution traces --- any other test case is simply thrown away.

Worth mentioning, CGPT is not meant to replace sophisticated manually-written
random generators, but rather to provide an adequate testing solution for early
stages of development.
%, and while a better test suite using manually-written generators is
%still under construction.

@@

In this work, we establish several aspects%
\footnote{Some of these were already identified by
  \citeauthor{lampropoulos2019coverage} in their original work.}
%
of the original CGPT approach by \citeauthor{lampropoulos2019coverage} that
leave considerable room for improvement (see \S \ref{sec:background}).
%
In particular:
%
\begin{inparaenum}
\item structure-preserving mutations, if not done carefully, can become
  superficial and rely heavily on randomness to produce diverse mutants,
\item the queuing mechanism can cause delays if interesting values are enqueued
  frequently and there is no way to prioritize them, and
\item using an heuristic to assign a certain ``mutation budget'' to each
  interesting test case (often referred to as a \emph{power schedule}) requires
  fine tuning and can be hard to generalize to different kinds of testing
  properties.
\end{inparaenum}
%
%% Moreover, we replicated the Information-Flow Control (IFC) stack machine case
%% study bundled with \fuzzchick and observed a surprising limitation:
%% %
%% after repeating each experiment 30 times, \fuzzchick could only find 5 out of
%% the 20 injected bugs with a 100\% efficacy.
%% %
%% %% the hardest one being found only around 13\% of the time after an hour of
%% %% testing.
%% %
%% Despite this, \citeauthor{lampropoulos2019coverage} show that \fuzzchick is far
%% better than using normal random testing coupled with na\"ive random generators
%% by comparing the mean-time-to-failure (MTTF) against \emph{QuickChick}.
%% %
%% However, we believe that a more meticulous evaluation ought to take failure to
%% find a counterexample as an important metric when comparing PBT tools, not just
%% mean time to failure --- when one is found.

@@

%% Validation
To validate our ideas, we use two case studies (described in detail in \S
\ref{sec:casestudies}): the Information-Flow Control (IFC) stack machine used by
\citeauthor{lampropoulos2019coverage} in their original work, and an existing
{Web-Assembly} engine of industrial strenght written in the strongly-typed
programming language Haskell.
%
In both case studies, we additionally compare the effect of the heuristics
implemented on top of the base testing loop of \mutagen.
%
In the IFC stack machine case study, we compare \mutagen against \fuzzchick, the
reference CGPT implementation developed in Coq \cite{lampropoulos2019coverage}.
%
Our results (\S \ref{sec:evaluation}) indicate that \mutagen outperforms
\fuzzchick both in terms of time to failure and failure rate.
%
On the other hand, our WebAssembly case study shows the performance of our tool
in a realistic scenario.
%
There, \mutagen is capable of reliably finding 15 planted bugs in the validator
and interpreter, as well as 3 previously unknown bugs that flew under the radar
of the existing unit test suite of this engine.
%
This case study also help us to compare the performance of our tool against a
more traditional RPBT approach that does not rely on code instrumentation.
%
All in all, our evaluation encompasses more than 600 hours of computing time and
suggests that \emph{testing mutants exhaustively together with our heuristics to
  escape scalability issues} can be an appealing technique for
finding bugs reliably without sacrificing speed.

@@

% ----------------------------------------
%% \vspace{-5pt}

@@

\noindent The type parameter \texttt{a} indicates that trees can be instantiated
using any other type as payload, so the value \texttt{Leaf True} has type
\texttt{Tree Bool}, whereas the value \texttt{Branch (Leaf 1) 2 (Leaf 3)} has
type \texttt{Tree Int}.
%
%% Then, we can define tree reflections using a simple recursive function that
%% pattern matches against the two possible constructors, inverting the order of
%% the subtrees whenever it encounters a branch:
%
%% \begin{verbatim}
%% mirror :: Tree a -> Tree a
%% mirror (Leaf x)       = Leaf x
%% mirror (Branch l x r) = Branch (mirror r) x (mirror l)
%% \end{verbatim}
%
%% Later, a reasonable requirement to assert for is that \texttt{mirror} must be
%% \emph{involutive}, i.e., reflecting a tree twice always yields the original
%% tree.
%% %
%
Then, if we assume existence of a predicate (function) \texttt{balanced} of type
\texttt{Tree a -> Bool} that asserts that a tree satisfies a certain notion of
balancedness, we can write testing properties to assert that the operations
defined over the \texttt{Tree} data type preserve this invariant.
%
For instance, the following testing property can be used to find bugs in the
implementation of a given \texttt{insert} function:

@@

\noindent This property asserts that, given an element \texttt{x} and a balanced
tree \texttt{t} as input, inserting \texttt{x} into \texttt{t} will produce a
balanced tree as output.
%
(The implementation details of \texttt{balanced} and \texttt{insert} are not
relevant for the point being made here.)
%
Moreover, the arrow operator (\texttt{==>}) indicates that the predicate
\texttt{balanced t} is a precondition of this property, and so attempting to
test it using an unbalanced tree as an input will result in the test case
getting \emph{discarded}.

The last missing piece is a random generator of trees.
%
%% In \quickcheck, this is commonly done via the type class mechanism
%% \cite{jones1997type}, instantiating the \texttt{Arbitrary} type class with a
%% random generator as the implementation of the overloaded \texttt{arbitrary}
%% operation:
We can write a na\"ive one for trees of integers (simplified to make it more accessible%
\footnote{In practice, this code will most likely be implemented using Haskell's
  type-class mechanism, hiding the complexity behind the size limit and the
  payload generator.}%
) as follows:

@@

The testing loop in CGPT uses two queues to store valid and discarded previously
executed test cases.
%
Enqueued test cases are stored along with a given mutation budget, that controls
how many times a given test case can be mutated before being finally thrown
away.
%
This budget is calculated using a heuristic derived from AFL's power schedule,
i.e., more budget to test cases that lead to shorter executions, or that
discover more parts of the code.
%
On each iteration, the algorithm selects the next test case by mutating the
first value on the queue of valid test cases. If that queue is empty, it then
selects the first one from the queue of discarded test cases.
%
If both queues are empty, CGPT generates a new random value from scratch.
%
The algorithm then run this test case and evaluates whether it was interesting
(i.e., it exercises a new path) based on its trace information.
%
If the test case does in fact discover a new path, it is enqueued at the end of
its corresponding queue, depending on whether it passed the property
precondition or was discarded.
%
%% (i.e., it failed the precondition).
%
This process alternates between generation and mutation until a bug is found or
it reaches the test limit.

@@

We evaluated the performance of \mutagen using two case studies.%
%
\footnote{An anonymized replication package \cite{anon_2021} is available for reviewing purposes.}
%
The first one is a simple abstract stack machine that enforces
the hyperproperty \emph{noninterference} \cite{goguen1982security} using runtime checks.
%
Its implementation was proven correct by \citeauthor{10.1145/2578855.2535839}
\citeyearpar{10.1145/2578855.2535839} in Coq, and subsequently degraded by
systematically introducing 20 bugs on its enforcing mechanism.
%
\citeauthor{lampropoulos2019coverage} used this case study to compare \fuzzchick
against RPBT using na\"ive- and hand-tuned random generators.
%
Here, we replicate their results and compare them against our tool.
%
Worth mentioning, we reimplemented this case study from scratch in Haskell in
order to run \mutagen on its test suite.

@@

\begin{table}[b]
\scriptsize
\begin{tabular}{|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Description} \\
\hline
1
& Validator
& Wrong if-then-else type validation on else branch \\
\hline
2
& Validator
& Wrong stack type validation \\
\hline
3
& Validator
& Removed function type mismatch assertion \\
\hline
4
& Validator
& Removed max memory instances assertion \\
\hline
5
& Validator
& Removed function index out-of-range assertion \\
\hline
6
& Validator
& Wrong type validation on \texttt{i64.eqz} instruction \\
\hline
7
& Validator
& Wrong type validation on \texttt{i32} binary operations \\
\hline
8
& Validator
& Removed memory index out-of-range assertion \\
\hline
9
& Validator
& Wrong type validation on \texttt{i64} constants \\
\hline
10
& Validator
& Removed alignment validation on \texttt{i32.load} instruction \\
\hline
11
& Interpreter
& Wrong interpretation of \texttt{i32.sub} instruction \\
\hline
12
& Interpreter
& Wrong interpretation of \texttt{i32.lt\_u} instruction \\
\hline
13
& Interpreter
& Wrong interpretation of \texttt{i32.shr\_u} instruction \\
\hline
14
& Interpreter
& Wrong local variable initialization \\
\hline
15
& Interpreter
& Wrong memory address casting on \texttt{i32.load8\_s} instruction \\
\hline
\end{tabular}
\caption{\label{table:wasm:injectedbugs}Bugs injected into
  \textit{haskell-wasm}.}
\vspace{-10pt}
\end{table}

>>>>

%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf, review]{acmart}

@@

\titlespacing*{\section}{0pt}{5pt}{2pt}
\titlespacing*{\subsection}{0pt}{5pt}{2pt}
\titlespacing*{\subsubsection}{0pt}{5pt}{2pt}

\titlespacing*{\paragraph}{10pt}{0pt}{10pt}

@@

%% These commands are for a JOURNAL article.
\acmConference[AST 2022]{3rd ACM/IEEE International Conference on Automation of Software Test}{May 21-22, 2022}{}

@@

\author{Agust\'in Mista}
\email{mista@chalmers.se}
\affiliation{%
  \institution{Chalmers University of Technology}
  \city{Gothenburg}
  \country{Sweden}
}

\author{Alejandro Russo}
\email{russo@chalmers.se}
\affiliation{%
  \institution{Chalmers University of Technology}
  \city{Gothenburg}
  \country{Sweden}
}

@@

\vspace{-5pt}

@@

%% In this light, CGPT is likely to preserve the semantic structure of the mutated
%% data, as mutations are applied directly at the data type level, e.g., the
%% program's AST in case of generating code.
%
Moreover, CGPT relies on execution traces to distinguish which executed test
cases were interesting and are therefore worth mutating --- a popular technique
known as \emph{coverage-guided fuzzing} and popularized by tools like \emph{AFL}
\cite{afl}.
%
Mutated test cases are considered interesting for mutation only when they
produce new execution traces --- any other test case is simply thrown away.
%
Worth mentioning, CGPT is not meant to replace sophisticated manually-written
random generators, but rather to provide an adequate testing solution for early
stages of development.
%, and while a better test suite using manually-written generators is
%still under construction.

@@

In this work, we establish several aspects of the original CGPT approach by
\citeauthor{lampropoulos2019coverage} that leave considerable room for
improvement (see \S \ref{sec:background}).
%
%
In particular:
%
\begin{inparaenum}
\item structure-preserving mutations, if not done carefully, can become
  superficial and rely heavily on randomness to produce diverse mutants,
\item the queuing mechanism can cause delays if interesting values are enqueued
  frequently and there is no way to prioritize them, and
\item using an heuristic to assign a certain ``mutation budget'' to each
  interesting test case (often referred to as a \emph{power schedule}) requires
  fine tuning and can be hard to generalize to different kinds of testing
  properties.
\end{inparaenum}
%
Notably, some of these limitations were already identified
\citeauthor{lampropoulos2019coverage} in their original work.
%
%% Moreover, we replicated the Information-Flow Control (IFC) stack machine case
%% study bundled with \fuzzchick and observed a surprising limitation:
%% %
%% after repeating each experiment 30 times, \fuzzchick could only find 5 out of
%% the 20 injected bugs with a 100\% efficacy.
%% %
%% %% the hardest one being found only around 13\% of the time after an hour of
%% %% testing.
%% %
%% Despite this, \citeauthor{lampropoulos2019coverage} show that \fuzzchick is far
%% better than using normal random testing coupled with na\"ive random generators
%% by comparing the mean-time-to-failure (MTTF) against \emph{QuickChick}.
%% %
%% However, we believe that a more meticulous evaluation ought to take failure to
%% find a counterexample as an important metric when comparing PBT tools, not just
%% mean time to failure --- when one is found.

@@

%% Validation
To validate our ideas, we use two case studies (described in detail in \S
\ref{sec:casestudies}): the Information-Flow Control (IFC) stack machine used by
\citeauthor{lampropoulos2019coverage} in their original work, and an existing
{Web-Assembly} engine of industrial strenght written in the strongly-typed
programming language Haskell.
%
In both case studies, we additionally compare the effect of the heuristics
implemented on top of the base testing loop of \mutagen.
%
In the IFC stack machine case study, we compare \mutagen against \fuzzchick, the
reference CGPT implementation developed in Coq \cite{lampropoulos2019coverage}.
%
Our results (\S \ref{sec:evaluation}) indicate that \mutagen outperforms
\fuzzchick both in terms of time to failure and failure rate.
%
On the other hand, our WebAssembly case study shows the performance of our tool
in a realistic scenario.
%
There, \mutagen is capable of reliably finding 15 planted bugs in the validator
and interpreter, as well as 3 previously unknown bugs that flew under the radar
of the existing unit test suite of this engine.
%
This case study also help us to compare the performance of our tool against a
more traditional RPBT approach that does not rely on code instrumentation.
%
All in all, our evaluation encompasses more than 600 hours of computing time and
suggests that \emph{testing mutants exhaustively together with our heuristics to
  escape scalability issues} can be an appealing technique for
finding bugs reliably without sacrificing speed.%
%
\footnote{A replication package \cite{anon_2021} is available for reviewing
  purposes.}

@@

% ----------------------------------------
%% \vspace{-10pt}

@@

\noindent The type parameter \texttt{a} indicates that trees can be instantiated
using any other type as payload, so the value \texttt{Leaf True} has type
\texttt{Tree Bool}, whereas the value \texttt{Branch (Leaf 1) 2 (Leaf 3)} has
type \texttt{Tree Int}.
%
%% Then, we can define tree reflections using a simple recursive function that
%% pattern matches against the two possible constructors, inverting the order of
%% the subtrees whenever it encounters a branch:
%
%% \begin{verbatim}
%% mirror :: Tree a -> Tree a
%% mirror (Leaf x)       = Leaf x
%% mirror (Branch l x r) = Branch (mirror r) x (mirror l)
%% \end{verbatim}
%
%% Later, a reasonable requirement to assert for is that \texttt{mirror} must be
%% \emph{involutive}, i.e., reflecting a tree twice always yields the original
%% tree.
%% %
%
Then, if we assume existence of a function \texttt{balanced} of type
\texttt{Tree a -> Bool} that asserts that a tree satisfies a certain notion of
balancedness, we can write testing properties to assert that the operations
defined over the \texttt{Tree} data type preserve this invariant.
%
For instance, the following testing property can be used to find bugs in the
implementation of a given \texttt{insert} function:

@@

\noindent This property asserts that, given an element \texttt{x} and a balanced
tree \texttt{t} as input, inserting \texttt{x} into \texttt{t} will produce a
balanced tree as output.
%
(The implementation details of \texttt{balanced} and \texttt{insert} are not
important here.)
%
Moreover, the arrow operator (\texttt{==>}) indicates that \texttt{balanced t}
is a precondition of this property, and so attempting to test it using an
unbalanced tree as input will result in the test case getting \emph{discarded}.

The last missing piece is a random generator of trees.
%
%% In \quickcheck, this is commonly done via the type class mechanism
%% \cite{jones1997type}, instantiating the \texttt{Arbitrary} type class with a
%% random generator as the implementation of the overloaded \texttt{arbitrary}
%% operation:
We can write a na\"ive generator for trees of integers (simplified here to make it
more accessible) as follows:
%% \footnote{In practice, this code will most likely be implemented using Haskell's
%%   type-class mechanism, hiding the complexity behind the size limit and the
%%   payload generator.}%

@@

The testing loop in CGPT uses two queues to store valid and discarded previously
executed test cases.
%
Enqueued test cases are stored along with a given mutation budget that controls
how many times a given test case can be mutated before being finally thrown
away.
%
This budget is calculated using a heuristic derived from AFL's power schedule,
i.e., more budget to test cases that lead to shorter executions, or that
discover more parts of the code.
%
On each iteration, the algorithm selects the next test case by mutating the
first value on the queue of valid test cases. If that queue is empty, it then
selects the first one from the queue of discarded test cases.
%
If both queues are empty, CGPT generates a new random value from scratch.
%
The algorithm then run this test case and evaluates whether it was interesting
(i.e., it exercises a new path) based on its trace information.
%
If the test case does in fact discover a new path, it is enqueued at the end of
its corresponding queue, depending on whether it passed the property
precondition or was discarded.
%
%% (i.e., it failed the precondition).
%
This process alternates between generation and mutation until a bug is found or
it reaches the test limit.

@@

We evaluated the performance of \mutagen using two case studies.%
%
The first one is a simple abstract stack machine that enforces
the hyperproperty \emph{noninterference} \cite{goguen1982security} using runtime checks.
%
Its implementation was proven correct by \citeauthor{10.1145/2578855.2535839}
\citeyearpar{10.1145/2578855.2535839} in Coq, and subsequently degraded by
systematically introducing 20 bugs on its enforcing mechanism.
%
\citeauthor{lampropoulos2019coverage} used this case study to compare \fuzzchick
against RPBT using na\"ive- and hand-tuned random generators.
%
Here, we replicate their results and compare them against our tool.
%
Worth mentioning, we reimplemented this case study from scratch in Haskell in
order to run \mutagen on its test suite.

@@

\begin{table}[b]
\scriptsize
\begin{tabular}{|c|c|l|}
\hline
\textbf{Id}
& \textbf{Subsystem}
& \textbf{Description} \\
\hline
1
& Validator
& Wrong if-then-else type validation on else branch \\
\hline
2
& Validator
& Wrong stack type validation \\
\hline
3
& Validator
& Removed function type mismatch assertion \\
\hline
4
& Validator
& Removed max memory instances assertion \\
\hline
5
& Validator
& Removed function index out-of-range assertion \\
\hline
6
& Validator
& Wrong type validation on \texttt{i64.eqz} instruction \\
\hline
7
& Validator
& Wrong type validation on \texttt{i32} binary operations \\
\hline
8
& Validator
& Removed memory index out-of-range assertion \\
\hline
9
& Validator
& Wrong type validation on \texttt{i64} constants \\
\hline
10
& Validator
& Removed alignment validation on \texttt{i32.load} instruction \\
\hline
11
& Interpreter
& Wrong interpretation of \texttt{i32.sub} instruction \\
\hline
12
& Interpreter
& Wrong interpretation of \texttt{i32.lt\_u} instruction \\
\hline
13
& Interpreter
& Wrong interpretation of \texttt{i32.shr\_u} instruction \\
\hline
14
& Interpreter
& Wrong local variable initialization \\
\hline
15
& Interpreter
& Wrong memory address casting on \texttt{i32.load8\_s} instruction \\
\hline
\end{tabular}
\caption{\label{table:wasm:injectedbugs}Bugs injected into
  \textit{haskell-wasm}.}
\vspace{-30pt}
\end{table}
